Formal verification of train route planning

### 2025-06-27 paper/main.tex
Schéma logiciel & TODOs & config PC lucas

\paragraph{Signals.}
%\todo{Est-ce qu'on met les signaux dans les agents ?}
At each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $\signals: \blocks \times \directions \to \{ \sigred, \siggreen\}$. The signal status is determined by the network configuration and explicit locking requests. Given some signal status $\signals$, a block $\bid{i}$ and a direction $\dirFmt{d}$, $\signalF{\bid{i}}{d} = \sigred$ means the signal guarding the exit of block $\bid{i}$ in direction $\dirFmt{d}$ is red, and we expect trains not leaving that block going in this direction\footnote{Our semantics, presented in Section~\ref{sec:model_semantics}, ensure such behaviour.}. We note $\setSignalsTo{\bid{\text{tgt}}}{c}$ the signal state $\signals$ except signals of block \bid{\text{tgt}}, which are turned to color $c$: 
$$(\setSignalsTo{\bid{\text{tgt}}}{c})(\bid{\text{from}}, \dirFmt{d}) = \begin{cases}
	c & \text{if }\bid{\text{from}} = \bid{\text{tgt}}\\
	\signalF{\bid{\text{from}}}{d} & \text{otherwise}
\end{cases}$$

@@

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers. A handler is a tuple $\tuple{\bid{i}, A}$, where $\bid{i}$ is the position the train occupies when triggering the event, and $A$ is the list of orders associated with that event. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.
Given a routing program $E$, we note $\handlerOf{E}{\tid{i}}$ the list of handlers of train $\tid{i}$, $\head{\handlerOf{E}{\tid{i}}} = \handlerOf{E}{\tid{i}}[1]$ the first handler of the list. Given a handler $h = \tuple{\bid{i}, A}$, let  $\actionsOf{h} \triangleq A$, $\blockOf{h} \triangleq \bid{i}$. When $h = \tuple{\bid{i}, a \cdot A}$ is the head of a sequence $H = h \cdot H'$,  $\nextAct{H} \triangleq \tuple{\bid{i}, A} \cdot H'$. %}{} \triangleq H'$ .  %(resp. \blockOf{h}) the list of action (resp. the block id) of $h$.
% We note $\nextAct{\handlerOf{E}{\tid{i}}}$ the routing program $E$ where the first event handler of train $\tid{i}$ is replaced with $\handler{\blockOf{\head{\handlerOf{E}{\tid{i}}}}}{\buftail{\actionsOf{\head{\handlerOf{E}{\tid{i}}}}}}$, i.e. where the first action of the first handler of train \tid{i} is removed.
% We note $\popHandlerHead{E}{\tid{i}}$ the routing program $E$ where the first event handler of train \tid{i} is removed, i.e. where $\handlerOf{E}{\tid{i}}$ is replaced by $\buftail{\handlerOf{E}{\tid{i}}}$.

@@

%\todo{Expliquer le temps discret et le mix grand pas/petit pas.}

@@

Then, the regulator lookup the first event on the buffer, and executes all the rules in the event handler associated with this event. All actions are treated in order, and when the list of remaining actions is empty, the handling terminates and the event is removed of the buffer. We introduce 6 reduction rules for the 4 possible kinds of orders (\incr{\bid{i}} and \wait{\bid{i}}{n} have two rules).

Rule \ruleRef{Turn} triggers when a \turn{\sid{i}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{\bufTrain} = \tid{i}
		\and
		\actionsOf{\head{\handlerOf{E}{\tid{i}}}} = \regConcat{\turn{\sid{i}}{s}}{A^\prime}
	}{
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
	\\\reduces
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{i}}}}{T}{W}}{\updateSwitch{\switches}{\sid{i}}{s}}{\signals}}
	}
\end{mathpar}

@@

The representation of the network topology is also represented within TLA+. It takes the shape of a function that is precomputed via a python script (see XXX%\todo{Référence au schéma}).

@@

The system is then represented by a formula:
\begin{verbatim}
    Spec == Init /\ [][Next]_variables /\ WF_variables(Next)
\end{verbatim}%Revoir la syntaxe au cas où / remplacer par le jolie tla2latex
where variables is the tuple of all variables. The intended meaning is that the initial configuration is specified in \mintinline{C}{Init} and steps follow the \mintinline{C}{Next} formula.  Without the weak fairness condition \mintinline{C}{WF_variables(Next)},  nothing forces a rule to be applied. As a consequence, an ongoing train could never reach its next block. The weak fairness condition prevents this incorrect simulation.

@@

\paragraph{Experiments}  With TLA+, we checked that any safe step is actually (not surprisingly!) reliable.  On our network, one finds 58 368 safe steps, all being reliable. Checking all the paths took XXX seconds on YYY. %\todo{XXX= nombre de seconde, YYY = décrire la machine (processeur, RAM, etc)}

@@

First,  for any train $\tid{i} \in \trains$, we append the orders of the train in $S_k$ to the corresponding ones in $S$. %let $P_0 = S.\Gamma[\tid{i}].P$ and $P_k = S_k.\Gamma[\tid{i}].P$, then set $\Gamma'[\tid{i}] = \Gamma[\tid{i}][ P \leftarrow P_0 \cdot P_k]$. 
Second, we proceed by updates of $E$ and $\sigma$. At first, we set $E' = E$ and $\sigma' = \sigma$. We begin to append to $E'$, for any train, all the handlers within $S_k$. To each train $\tid{i}$, we had for its first step (within $X_k$, seen within the composition): an $\auth$ order in $E'[\tid]$.

@@

\paragraph{Optimisations. } We apply some optimisations on the orders. For instance, we remove orders that do not alternate: $\su{d}{\vec{b}} \cdot \su{d}{\vec{b}} \cdot P$ by $\su{d}{\vec{b} \cdot \vec{b'}} \cdot P$. In the same way,  we remove empty lists, singleton lists: if $Len(\vec{b}) \leq 1$, we replace $\su{d}{\vec{b}} \cdot P$ by $P$. We have few more heuristics that we do not show here due to lack of space.

\paragraph{Experiments}  Parler des expériences

@@

\paragraph{Contributions.}
This paper introduces three main contributions. First we presented a formal model for railway network. Our model simplifies multiple elements from real life networks, but nonetheless keeps non-trivial elements, such as trackworks and signals. Contrary to existing models, ours makes very few assumptions about the network: tracks are two ways and %\todo{...}

@@

Our implementation, including both the TLA model and the Python composition is available on...%\todo{Donner le lien vers github ? }.

@@

%\todo{On peut avoir des routages de n'importe où, donc on peut voir pour interfacer avec d'autres approches ?}

>>>>

\paragraph{Signals.}
At each end of the blocks, there is a signal whose state is \sigred (red) or \siggreen (green). In other words, the status of signals is described by a function $\signals: \blocks \times \directions \to \{ \sigred, \siggreen\}$. The signal status is determined by the network configuration and explicit locking requests. Given some signal status $\signals$, a block $\bid{i}$ and a direction $\dirFmt{d}$, $\signalF{\bid{i}}{d} = \sigred$ means the signal guarding the exit of block $\bid{i}$ in direction $\dirFmt{d}$ is red, and we expect trains not leaving that block going in this direction\footnote{Our semantics, presented in Section~\ref{sec:model_semantics}, ensure such behaviour.}. We note $\setSignalsTo{\bid{\text{tgt}}}{c}$ the signal state $\signals$ except signals of block \bid{\text{tgt}}, which are turned to color $c$: 
$$(\setSignalsTo{\bid{\text{tgt}}}{c})(\bid{\text{from}}, \dirFmt{d}) = \begin{cases}
	c & \text{if }\bid{\text{from}} = \bid{\text{tgt}}\\
	\signalF{\bid{\text{from}}}{d} & \text{otherwise}
\end{cases}$$

@@

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers. A handler is a tuple $\tuple{\bid{i}, A}$, where $\bid{i}$ is the position the train occupies when triggering the event, and $A$ is the list of orders associated with that event. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.
Given a routing program $E$, we note $\handlerOf{E}{\tid{i}}$ the list of handlers of train $\tid{i}$, $\head{\handlerOf{E}{\tid{i}}} = \handlerOf{E}{\tid{i}}[1]$ the first handler of the list. Given a handler $h = \tuple{\bid{i}, a \cdot A}$, let  $\actionsOf{h} \triangleq a$, $\blockOf{h} \triangleq \bid{i}$ and $\nextAct{h} \triangleq \tuple{\bid{i}, A}$.  %(resp. \blockOf{h}) the list of action (resp. the block id) of $h$.
% We note $\nextAct{\handlerOf{E}{\tid{i}}}$ the routing program $E$ where the first event handler of train $\tid{i}$ is replaced with $\handler{\blockOf{\head{\handlerOf{E}{\tid{i}}}}}{\buftail{\actionsOf{\head{\handlerOf{E}{\tid{i}}}}}}$, i.e. where the first action of the first handler of train \tid{i} is removed.
 We note $\popHandlerHead{E}{\tid{i}}$ the routing program $E$ where the first event handler of train \tid{i} is removed, i.e. where $\handlerOf{E}{\tid{i}}$ is replaced by $\buftail{\handlerOf{E}{\tid{i}}}$.

@@

@@

Then, the regulator lookup the first event on the buffer, and executes all the rules in the event handler associated with this event. All actions are treated in order, and when the list of remaining actions is empty, the handling terminates and the event is removed of the buffer. We introduce 6 reduction rules for the 4 possible kinds of orders (\incr{\bid{i}} and \wait{\bid{i}}{i} have two rules).

Rule \ruleRef{Turn} triggers when a \turn{\sid{i}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{\bufTrain} = \tid{i}
		\and
		\head{\handlerOf{E}{\tid{i}}} = \handler{\bid{i}}{\regConcat{\turn{\sid{i}}{s}}{A^\prime}}
	}{
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{E}{T}{W}}{\switches}{\signals}}
	\\\reduces
	\redTuple{\guardR}{\bufTrain}{\bufSig}{\stateTuple{\trainSeq}{\regTuple{\nextAct{\handlerOf{E}{\tid{i}}}}{T}{W}}{\updateSwitch{\switches}{\sid{i}}{s}}{\signals}}
	}
\end{mathpar}

@@

The representation of the network topology is also represented within TLA+. It takes the shape of a function that is precomputed via a python script (see \texttt{to\_tla.py} in Fig.~\ref{fig:toolchain} on page~\pageref{fig:toolchain}).

@@

The system is then represented by a formula:
\begin{verbatim}
    Spec == Init /\  [][Next]_variables
\end{verbatim}%Revoir la syntaxe au cas où / remplacer par le jolie tla2latex
where variables is the tuple of all variables. The intended meaning is that the initial configuration is specified in \mintinline{C}{Init} and steps follow the \mintinline{C}{Next} formula.  Actually, presently, in our formal semantics, nothing forces a rule to be applied. As a consequence, an ongoing train may never reach its next block. Thanks to TLA+, we can ensure that a rule will eventually be applied (if it can be). Those are known as weak fairness conditions. Again, we specify it as follows. \todo{WF condition?}

@@

\paragraph{Experiments}  With TLA+, we checked that any safe step is actually (not surprisingly!) reliable.  On our network, one finds 58 368 safe steps, all being reliable. Checking all the paths took approximately 30 hours on an Linux desktop machine with an Intel i5-4590 (4-cores at 3.7GHz) and 8GB of memory.

@@

First,  for any train $\tid{i} \in \trains$, we append the orders of the train in $S_k$ to the corresponding ones in $S$. %let $P_0 = S.\Gamma[\tid{i}].P$ and $P_k = S_k.\Gamma[\tid{i}].P$, then set $\Gamma'[\tid{i}] = \Gamma[\tid{i}][ P \leftarrow P_0 \cdot P_k]$. 
Second, we proceed by updates of $E$ and $\sigma$. At first, we set $E' = E$ and $\sigma' = \sigma$. We begin to append to $E'$, for any train, all the handlers within $S_k$.

@@

\subsection{Simplification}

\subsection{Experiments}

\begin{figure}
	\begin{tikzpicture}
		\tikzstyle{programme} = [
		draw=black,
		]
		\tikzstyle{fichier} = [
		rounded corners,
		draw=blue,
		]
		\tikzstyle{generated} = [
		fill=blue!10,
		]
		\node[programme] (main) {main.py};
		\node[fichier, below=of main] (json) {safe\_steps.json};
		\node[programme, right = of main] (compose) {compose.py};
		\node[programme, right=of compose] (to_tla) {to\_tla.py};
		\node[fichier, below=of to_tla, generated] (compo) {route.tla};
		\node[programme, right=of compo] (tlc) {tlc};
		\node[fichier, above=of tlc] (model){model.tla};	
		\node[fichier, above right=of tlc, generated] (out) {out.log};
		\node[fichier, right=of tlc, generated] (err) {errors.log};

		\draw[stealth-] (json) to node[sloped, anchor=south] {uses} (main);
		\draw[-stealth] (main) to node[sloped, anchor=south] {calls} (compose);
		\draw[-stealth] (main) to[bend right] node[sloped, anchor=south] {calls} (to_tla);
		\draw[-stealth] (to_tla) to node[anchor=west] {creates} (compo);
		\draw[-stealth] (tlc) to node[sloped, anchor=south] {reads} (compo);
		\draw[-stealth] (tlc) to node[sloped, anchor=south] {reads} (model);
		\draw[-stealth] (tlc) to node[sloped, anchor=north] {creates} (err);
		\draw[-stealth] (tlc) to node[sloped, anchor=south] {creates} (out);
	\end{tikzpicture}
	\caption{Overview of the toolchain. Black rectangles are programs. Blue rounded rectangles are files, included generated files filled in blue. The main script uses a json file with safe steps (c.f. Section~\ref{sec:composition}) to produce composed routes (calling the script \texttt{compose.py}). Routes are then converted in TLA with \texttt{to\_tla.py}, producing a file \texttt{route.tla}. \texttt{tlc}, a TLA model checker, then uses the route and the model to check for correctness, producing log files.}
	\label{fig:toolchain}
\end{figure}

Parler des expériences

@@

\paragraph{Contributions.}
This paper introduces three main contributions. First we presented a formal model for railway network. Our model simplifies multiple elements from real life networks, but nonetheless keeps non-trivial elements, such as trackworks and signals.

@@

Our implementation, including both the TLA model and the Python composition is available on...\todo{Donner le lien vers github ? }.

@@
